---
title: "A More Comprehensive Application to Binary Classification Using Logistic Regression"
author: "zhengliang"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::knit_hooks$set(purl = knitr::hook_purl)
```

## 1 Binary Classification Using Other Dataset With Out-of-Sample Evaluation

In the first project of Binary Classification Using Logistic Regression Models (click [here](https://tinyurl.com/c276mvp7) to access it), I explored the basics of logistic regression using the **`glm`** function on the `Pima Indians Diabetes` dataset. However, real world data sets are likely to have more attributes/variables which we might not be able to select variables simply based off p-values to indicate significance of variables. Another possible problem was the usefulness of the model to predict outcomes on out-of-sample data.

I sought to address these problems in this section by applying more real-world techniques and comparing the predictive power of the different models. However, one should remember that the purpose of machine learning and statistical modelling are different. Machine learning usually aims for best prediction while statistical modelling might be more towards the inference of explanatory variables to the response variable.

## 2 Packages Required

```{r load packages, message=FALSE}
# Use install.packages("packagename") if the packages are not installed in R
library(car) 
library(caret) # For building ML models
library(corrplot) # For correlation plot
library(tidyverse) # For ggplot and dplyr packages
library(mlbench) # For data sets
library(lmtest) # For statistical tests
library(glmnet) # For LASSO logistic regression
library(gridExtra) # Plot and arrange multiple graphs in a single pane
library(ROCR) # For determining cutoff-probability using ROC
library(readxl) # Fore reading Excel files
library(stargazer) # For neat tables and outputs, where possibles
```

## 3 Dataset and Exploratory Data Analysis

The dataset that I used is called [`default of credit card clients`](https://archive.ics.uci.edu/ml/datasets/default+of+credit+card+clients), which can be found in the UCI Machine Learning Repository. It contains customer credit default data from Taiwan in 2005. Other than the binary classification of default and non-default, the probability of default may also be an important result. The file downloaded is an Excel file and to load it into R, we need the **`readxl`** package.

```{r load dataset from wd, message=FALSE}
# This step loads the file from the working directory. If unsure of wd, use 
# getwd() to check the location of the wd and setwd() to set the wd where the file is saved.

default.data <- data.frame(read_excel("credit card default.xls"))
```

Let us take a peek at `default.data`.

```{r head tail}
# Show only the first 10 columns as there are too many variables

stargazer(rbind(head(default.data[,1:10]), tail(default.data[,1:10])),
          type = "text",
          title = "First and Last 6 Observations in Default Data",
          summary = F)

colnames(default.data)
```

From the **`head`** function, we saw that the first row should be the column names of the data frame and the first column should be the row names. There are many variables, 24 to be exact, with 1 dependent variable (Y, the last column) and 23 independent variables (X1 to X23). I would start by adjusting the column and row names.

```{r adjust column names}
colnames(default.data) <- default.data[1,]

default.data <- default.data[-1,]
```

```{r adjust row names}
# Remove the first column since ID matches the auto-assigned row numbers.
default.data <- default.data[,-1]

rownames(default.data) <- NULL # Reset the row numbers

stargazer(head(default.data[,1:10]),
          type = "text",
          title = "First 6 Observations in Default Data, Adjusted",
          summary = F)
```

Next, I would change the some of the column names to be something more descriptive but it is optional.

```{r renaming columns}
newnames <- c("PAY_SEP","PAY_AUG","PAY_JUL", "PAY_JUN", "PAY_MAY", "PAY_APR",
              "BILLAMT_SEP","BILLAMT_AUG","BILLAMT_JUL", "BILLAMT_JUN", "BILLAMT_MAY", "BILLAMT_APR",
              "PAYAMT_SEP","PAYAMT_AUG","PAYAMT_JUL", "PAYAMT_JUN", "PAYAMT_MAY", "PAYAMT_APR")

oldnames <-names(default.data %>%
                   dplyr::select(grep("PAY" , names(default.data)), 
                                 grep("BILL", names(default.data)))) 

colnames(default.data)[colnames(default.data) %in% oldnames] <- newnames

default.data <- default.data %>%
  rename(default = `default payment next month`)

colnames(default.data)
```

I have listed the description of the variables in the table below, which can be found in the UCI Machine Learning Repository where the data set was downloaded from.

| Variable                  | Description                                                                                                                                        |
|------------------------|-----------------------------------------------|
| LIMIT_BAL                 | Amount of the given credit, including both the individual consumer credit and his/her family (supplementary) credit                                |
| SEX                       | Gender. Male = 1, Female = 2                                                                                                                       |
| EDUCATION                 | Graduate School = 1, University = 2, High School = 3, Others = 4                                                                                   |
| MARRIAGE                  | Married = 1, Single = 2, Others = 3                                                                                                                |
| AGE                       | In years                                                                                                                                           |
| PAY_SEP - PAY_APR         | History of past payment from April to September 2005. Pay duly = -1, Delay one month = 1, Delay two months = 2, ..., Delay nine months or more = 9 |
| BILLAMT_SEP - BILLAMT_APR | Amount of bill statement from April to September 2005.                                                                                             |
| PAYAMT_SEP - PAYAMT_APR   | Amount of previous payment.                                                                                                                        |
| default                   | Response variable. Yes = 1, No = 0.                                                                                                                |

Now, we can begin exploring the data set through an Exploratory Data Analysis.

```{r structure}
str(default.data)
```

We can see that all the data values has been loaded as characters, and we need to change this to their appropriate types (numeric, factors, etc.).

```{r changing value types}
# Changing variables PAYAMT, BILLAMT, AGE and LIMIT_BAL to numeric
default.data <- default.data %>% 
  mutate(
    across(
      c(grep("PAYAMT", names(default.data)), 
        grep("BILLAMT", names(default.data)), 
        "AGE", "LIMIT_BAL"), 
      as.numeric))

# Changing variables SEX, EDUCATION, MARRIAGE and default to factor
default.data <- default.data %>%
  mutate(
    across(
      c("SEX", "EDUCATION", "MARRIAGE", "default",
        grep("PAY_", names(default.data))),
      as.factor))
```

```{r summary}
summary(default.data)
```

We can see that there are some variables that stood out. For example, in the variables *`EDUCATION`* and *`MARRIAGE`*, there are other classes that were not known beforehand. The variables *`PAY_`* has values of -2 and 0, which were also not indicated in the description of the data. This could probably have meant prepayments and paying within the month respectively, and since they could provide information about how paying habits affect whether a person defaulted, we should not remove these observations but keep in mind that there is ambiguity in the data set.

Let us clean the data for variables *`EDUCATION`* and *`MARRIAGE`*. Since they have each have a class called `Others`, we can consider replacing the unknown classes with this class.

```{r replace unknown class with others}
# For EDUCATION
levels(default.data$EDUCATION)[levels(default.data$EDUCATION) %in% c("0", "5", "6")] <- "4"

# For MARRIAGE
levels(default.data$MARRIAGE)[levels(default.data$MARRIAGE) == "0"] <- "3"

default.data %>% select(where(is.factor)) %>% str()
```

After the cleaning of the *`EDUCATION`* and *`MARRIAGE`* variables, we can see that only 4 and 3 levels are left, from the original 7 and 4 respectively. The *`PAY_`* variables have different number of levels, which we can check when we further analyze the variables. We can plot out the variables in a univariate and multivariate plots to visualize how they are distributed and correlate with one another.

### 3.1 Univariate Analysis

We can first take a look at the distribution of the *`default`* variable.

```{r dist of default, fig.align="center"}
ggplot(data = default.data, aes(x = default, fill = default)) +
  geom_bar()
```

It can be seen that this is an imbalanced data set, with much higher number of non-default observations compared to default observations. We would need to correct these using under-sampling, over-sampling or SMOTE methods, and we can compare their in-sample and out-of-sample predictions against uncorrected data. 

I have included plots of the other factor variables below. Using the package **`gridExtra`** allows us to arrange a number of plots into a single page.

```{r dist of sex_educ_marr_age, message=FALSE, fig.align="center"}
# For categorical and discrete variables
sex <- ggplot(data = default.data) +
  geom_bar(aes(x = SEX, fill = default)) +
  theme(legend.title = element_text(size = 10),
        legend.key.size = unit(0.5, "cm"),
        axis.title.x = element_text(size = 10))

education <- ggplot(data = default.data) + 
  geom_bar(aes(x = EDUCATION, fill = default)) +
  theme(legend.title = element_text(size = 10),
        legend.key.size = unit(0.5, "cm"),
        axis.title.x = element_text(size = 10))

marriage <- ggplot(data = default.data) +
  geom_bar(aes(x = MARRIAGE, fill = default)) +
  theme(legend.title = element_text(size = 10),
        legend.key.size = unit(0.5, "cm"),
        axis.title.x = element_text(size = 10))

age <- ggplot(data = default.data) +
  geom_histogram(aes(x = AGE, fill = default)) +
  theme(legend.title = element_text(size = 10),
        legend.key.size = unit(0.5, "cm"),
        axis.title.x = element_text(size = 10))

gridExtra::grid.arrange(sex, education, marriage, age, ncol = 2, nrow = 2)
```

For the *`PAY_`* variables, I have used a contingency table as there are several of these variables with multiple levels, making plots difficult to see.

```{r contingency table}
colno <- grep("PAY_", colnames(default.data))

cont_tab <- list()

for (i in colno) {
  temp_tab <- table(default.data$default, default.data[,i])
  
  name <- colnames(default.data)[i]
  
  cont_tab[[name]] <- temp_tab
}

cont_tab
```

The contingency table shows an interesting observation, where *`PAY_MAY`* and *`PAY_APR`* do not have the factor class of "1", explaining the difference in the number of levels in the *`PAY_`* variables. However, there are too many levels and we can see that typically levels -2, -1 and 0 have the most observations, followed by 1, 2 and 3. Levels 4 to 8 appears to have few observations. We may want to reclassify the levels into just these 3 groups as more variables generally affects variance of our estimators and the reliability of any statistical tests performed.

From the plots, it is hard to determine if the *`default`* variable is correlated with gender, education, marriage and age. This is because while we see higher default in certain groups, those groups also had more observations which would allow us to find more people who defaulted.

```{r dist billamt and payamt, fig.align="center"}
# For continuous variables
billamt <- default.data %>%
  select(grep("BILLAMT", names(default.data)))

payamt <- default.data %>%
  select(grep("PAYAMT", names(default.data)))

par(mfrow = c(1, 6), mar = c(1, 3, 3, 2))
for (i in 1:6) {
  boxplot(billamt[,i], main = names(billamt)[i])
}

par(mfrow = c(1, 6), mar = c(1, 3, 3, 2))
for (i in 1:6) {
  boxplot(payamt[,i], main = names(payamt)[i])
}
```

From the boxplots, we can tell that the continuous variables are not normally distributed as they have very long tails (a lot of outlier points).

### 3.2 Multivariate Analysis

Let us visualize the correlation between the numeric variables in `default.data`.

```{r corrplot, fig.align="center"}
# Calculate the correlation between continuous variables
correl <- default.data %>% 
  select(where(is.numeric)) %>%
  cor(method = "spearman")

# Plot the correlation in a heatmap
corrplot(corr = correl,
         method = "color",
         type = "lower",
         main = "Correlation Matrix of Continuous Variables in Default Data",
         mar = c(0, 0, 1.5, 0))
```

We can see that the correlation between *`BILLAMT_`* variables is very high, which is likely due to the time component involved in these variables. We could expect that if the bill statement in April is high, then the bill statement in May would likely be high as well. However, it seemed that the correlation between the *`PAYAMT_`* is rather moderate. For the correlation between *`BILLAMT_`* and *`PAYAMT_`*, there seems to be a pattern of relatively higher correlation on the diagonals, i.e. between *`BILLAMT_AUG`* and *`PAYAMT_SEP`* or *`BILLAMT_JUL`* and *`PAYAMT_AUG`*. Due to the high correlation between *`BILLAMT_`* variables, I may need to create new features to eliminate the problem multicollinearity.

### 3.3 Creating New Variables

One variable I created is the deviation of the maximum utilization from the mean utilization over the period of April to September. Utilization would be obtained from an observation's *`BILLAMT_`* variables divided by its *`LIMIT_BAL`*. I created this variable as I believed that a higher deviation over the 6-month period may translate to a higher probability of default.

```{r avg and max util}
# Calculate average using rowMeans function
avg_util <- round((rowMeans(billamt) / default.data$LIMIT_BAL)*100, 2)

# Calculate max of each row using apply function and calling max function
max_util <- round((apply(billamt, 1, max) / default.data$LIMIT_BAL)*100, 2)

# Calculate deviation of maximum utilization from mean utilization
dev_util <- max_util - avg_util
```

The third variable I created is the average payment over the period April to September and the fourth is the difference between the average bill amount and average payment between April to September.

```{r avg pmt}
avg_pmt <- round(rowMeans(payamt), 2)

pmt_bill_diff <- round(rowMeans(billamt) - rowMeans(payamt),2)
```

I have bound the variables of interest into a new data frame called `new_data`.

```{r bind variables}
new_data <- data.frame(default.data[,1:11],
                       dev_util,
                       avg_pmt,
                       pmt_bill_diff,
                       default = default.data$default)

stargazer(rbind(head(new_data), tail(new_data)), 
          type = "text", 
          title = "First and Last 6 Observations in New Data", 
          summary = F,
          digits = 2)
```

The variables *`PAY_`* has too many classes, and there would be a need to reduce them. As seen in the contingency table produced in Section 3.1 for the *`PAY_`* variables, I have reduced the original 11 classes into 3. 0 = those who paid duly/early , 1 = those who delayed payment for one to three months, and 2 = those who delayed for more than three months.

```{r changing classes}
# Set level 0 for those who were originally given levels -2, -1 and 0

for (i in colno) {
  levels(new_data[,i])[levels(new_data[,i]) %in% c("-2", "-1", "0")] <- "0"
  
  levels(new_data[,i])[levels(new_data[,i]) %in% c("1", "2", "3")] <- "1"
  
  levels(new_data[,i])[!levels(new_data[,i]) %in% c("-2", "-1", "0", "1", "2", "3")] <- "2"
}

summary(new_data)
```

I calculated the correlation and plotted the correlation matrix to see if the numeric variables in `new_data` data frame are highly correlated. It can be seen that none of the variables have a correlation of more than 0.7, so multicollinearity might not be a problem.

```{r correl new var, fig.align="center"}
correl <- new_data %>% 
  select(where(is.numeric)) %>%
  cor(method = "spearman")

corrplot(corr = correl,
         method = "color",
         type = "lower",
         addCoef.col = "black",
         main = "Correlation Matrix of Continuous Variables in New Data",
         mar = c(0, 0, 1.5, 0))
```

## 4 Building the Models

Now that I have settled on the variables to be used for the logistic regression model, I can start to built the different models that are commonly used. Before modelling, I would split the data into training and testing sets in a 80:20 split.

```{r split data}
# Split data into 80% training and 20% testing sets
set.seed(1)
validation <- createDataPartition(new_data$default, p = 0.80, list = F)

traindata <- new_data[validation,]

testdata <- new_data[-validation,]
```

### 4.1 Full Model

I would start with building a full model that contains all the explanatory variables from the new data set created in Section 3.3, which can be used as a benchmark for evaluation.

```{r full model}
full.model <- glm(default == "1" ~ .,    # 1 refers to those who default
                  data = traindata,
                  family = binomial(link = "logit"))

stargazer(full.model, 
          type = "text", 
          title = "Regression Output of Full Model")
```

I used the model to calculate the variance inflation factor of each variable to test for multicollinearity. It can be seen that none of the variables has a large VIF value, so multicollinearity may not be a problem.

```{r vif2}
car::vif(full.model)
```

### 4.2 Restricted Model

From the regression results of the `full.model`, both of the *`MARRIAGE`* factor variables seemed to be statistically insignificant, so I removed them while keeping the rest of the variables in the restricted model.

```{r rest model wo marriage}
# Remove marriage variables
rest.model <- glm(default == "1" ~ .-MARRIAGE,
                   data = traindata,
                   family = binomial(link = "logit"))

stargazer(rest.model, 
          type = "text", 
          title = "Regression Output of Restricted Model")
```

I did a likelihood ratio test and found that the full model is preferred since the p-values showed strong evidence against the null hypothesis.

```{r lrtest2}
lmtest::lrtest(full.model, rest.model) # Result: Use full model
```

### 4.3 Checking for Influential Outliers in the Full Model

I checked for highly influential observations using Cook's Distance and outliers using standardized and studentized residuals.

```{r cooksd}
# Influence using Cook's Distance
plot(full.model, which = 4)

head(sort(cooks.distance(full.model), decreasing = T))
```

```{r standardized resid}
# Outliers with standardized/studentized residuals
std.resid <- rstandard(full.model)

val <- abs(std.resid) > 3

table(val)

val[val == "TRUE"] # Can be used to further investigate the observations
```

```{r studentized resid2}
car::outlierTest(full.model) # No outliers detected by studentized residuals
```

There was 1 influential outlier detected by the standardized residuals, which could be further investigated if necessary.

```{r looking at influential outlier}
# Use this command to get data on the influential outlier(s)
traindata[val == "TRUE",]
```

### 4.4 Checking Linearity Assumption of Full Model

To check the linearity assumption, I first found the fitted values (or the probability) using the full model. Next, I calculated the log-odds (or logit), and finally plotted the log-odds against all continuous variables in my model.

```{r fitvalues and logit}
prob.fullmod <- fitted.values(full.model)

logit = log(prob.fullmod/(1-prob.fullmod))
```

```{r plot, fig.align="center", message=FALSE, warning=FALSE}
p1 <- ggplot(data = traindata, aes(x = logit, y = LIMIT_BAL)) +
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "loess")

p2 <- ggplot(data = traindata, aes(x = logit, y = AGE)) +
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "loess")

p3 <- ggplot(data = traindata, aes(x = logit, y = dev_util)) +
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "loess")

p4 <- ggplot(data = traindata, aes(x = logit, y = avg_pmt)) +
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "loess")

p5 <- ggplot(data = traindata, aes(x = logit, y = pmt_bill_diff)) +
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "loess")

gridExtra::grid.arrange(p1, p2, p3, p4, p5, ncol = 2, nrow = 3)
```

It would seem that the continuous variables are mostly linear to the log-odds, maybe other than for the *`LIMIT_BAL`* variable.

## 5 Stepwise Model Using Backward Elimination

Most of the times, when there are too many variables even after manual selection by understanding business and data interactions, we may have to use some techniques to help us select variables/features. One of these techniques is the use of step-wise models, using either forward selection or backward elimination or a combination of both. 

These techniques selects variables that are usually statistically significant or meet certain requirements, such as lowest AIC or BIC. However, there are many arguments online and from research articles against using such techniques. But it does provide a quick and easy way to get a model from multiple explanatory variables and we can use it as a benchmark. I would use the backward elimination method, but the code is similar for the methods.

### 5.1 Creating the Model

Backward elimination starts off with a model that contains all the explanatory variables and removes them one at a time to reduce AIC, with the variable that causes the largest reduction to be removed first. Before that, let us create a null model that does not use any explanatory variables.

```{r null model}
# Create a null model that predicts using grand mean
null.model <- glm(default == "1" ~ 1,
                  data = traindata,
                  family = binomial(link = "logit"))

stargazer(null.model,
          type = "text",
          title = "Regression Output of Null Model")
```

```{r bwd model}
bwd.model <- step(full.model,
                  scope = list(lower = null.model, upper = full.model),
                  direction = "backward", 
                  k = 2,
                  trace = T)
# Indicate scope argument to define range of models to examine
# Indicate direction = "backward" for backward elimination, "forward" for forward selection
# and "both" for stepwise regression
# Indicate k = 2 for AIC, k = log(n) for BIC
# Indicate trace = F if want to suppress the running of the step by step regression

stargazer(bwd.model,
          type = "text",
          title = "Regression Output of Backward Elimination Model")
```

Based on the backward elimination approach, the variable *`dev_util`* has been removed to achieve the lowest AIC value, although the difference is very small.

## 6 LASSO Logistic Model

Another method of selecting explanatory variables is the LASSO regression, which shrinks the coefficients of variables that are not important to zero. This can be done in R using the **`glmnet`** package.

### 6.1 Converting Data Frame to Matrix

Before we can run the lasso logistic regression, the data frame has to be converted into a matrix for **`glmnet`** to run. We also need to convert categorical variables into suitable dummy variables. All this can be done using the **`model.matrix`** function.

```{r model.matrix}
x_var <- model.matrix(default ~ ., traindata)[,-1] # Remove intercept column

colnames(x_var) # Does not include the y variable

y_var <- ifelse(traindata$default == "1", 1, 0)
```

### 6.2 Calculating lambda

For LASSO regression, there is a need to specify the lambda value, which is the amount of the penalty added to the log-likelihood function. The optimal lambda value can be obtained using the **`cv.glmnet`** function. I have chosen to use the `lambda.1se` value which would give a simpler model, but `lambda.min` can also be chosen to obtain a more complex model.

```{r lambda}
lasso.lambda <- glmnet::cv.glmnet(x = x_var, 
                                  y = y_var, 
                                  alpha = 1,
                                  family = "binomial",
                                  type.measure = "class")

# alpha = 1 for lasso, family = "binomial" for logistic regression
# type.measure = "class" to find lambda that optimizes mis-classification error, 
# "mse" can also be used to optimize mean squared error
```

```{r optimal lambda}
# optimal value of lambda
opt.lambda <- lasso.lambda$lambda.min

opt.lambda
```

### 6.3 Fitting the LASSO Logistic Model

Now that the lambda value has been obtained and stored, the LASSO logit regression model can be fitted. Unfortunately, the **`summary`** function does not show the coefficients like the models fitted using **`glm`**. To extract the intercept and coefficients, add `$a0` and `$beta` to the model name.

```{r lasso model}
lasso.model <- glmnet(x_var,
                      y_var,
                      alpha = 1,
                      family = "binomial",
                      lambda = opt.lambda)

lasso.model$a0; lasso.model$beta # intercept and coefficients
```

### 6.4 Evaluating the LASSO Logistic Model

```{r ROC lasso is}
res_lassomodel <- predict(lasso.model, x_var, type = "response")

ROCpred <- ROCR::prediction(res_lassomodel, factor(y_var))

ROCperf <- ROCR::performance(ROCpred, "tpr", "fpr")

plot(ROCperf, 
     colorize = T, 
     print.cutoffs.at = seq(0, by = 0.01),
     text.adj = c(-0.2,1))
```

```{r conMat lasso is}
## Use 0.17 as threshold
pred.lasso <- as.factor(ifelse(res_lassomodel > 0.17, "1", "0"))

caret::confusionMatrix(factor(y_var),
                       pred.lasso,
                       dnn = c("Actual", "Predicted"))
```

The LASSO logistic regression model has led to a higher accuracy than the full model and the stepwise model. However, there is an increase in the misclassification of actual defaults as non-defaults.

```{r convert testdata to matrix}
xtest <- model.matrix(default ~ ., testdata)[,-1] # Remove intercept column
ytest <- ifelse(testdata$default == "1", 1, 0)
```

```{r conMat lasso oos}
res_lassomodel <- predict(lasso.model, xtest, type = "response")

pred.lasso <- as.factor(ifelse(res_lassomodel > 0.17, "1", "0"))

caret::confusionMatrix(factor(ytest),
                       pred.lasso,
                       dnn = c("Actual", "Predicted"))
```

With the out-of-sample data, banks might refuse giving credit to 1087 borrowers who are unlikely to default and give credit to 477 borrowers who are likely to default by using this model.

## 7 Evaluation of Models In-Sample

### 7.1 Full Model

```{r prob using full model}
# Returns the probability of default
res_fullmodel <- predict(full.model, traindata, type = "response")
```

```{r ROC and plot}
# First argument: predicted probability, Second argument: observed response
ROCpred <- ROCR::prediction(res_fullmodel, traindata$default)

# Performance based on true positive rate and false positive rate
ROCperf <- ROCR::performance(ROCpred, "tpr", "fpr")

# Plot ROC and indicate threshold values
plot(ROCperf, 
     colorize = T, 
     print.cutoffs.at = seq(0, by = 0.01),
     text.adj = c(-0.2,1))
```

```{r conMat fullmodel is}
# Use 0.17 as threshold
pred.fullmodel <- as.factor(ifelse(res_fullmodel > 0.17, "1", "0"))

caret::confusionMatrix(traindata$default, 
                       pred.fullmodel, 
                       dnn = c("Actual", "Predicted"))
```

Using `fullmodel` on the in-sample data, 12746 observations were correctly predicted as non-default and 3638 were correctly predicted as default. Since non-default is the positive class, there were 5946 false negative and 1671 false positive. In other words, the model incorrectly predicted 1671 actual defaults as non-defaults. This could have serious implications to a financial institution depending on the size of the credit. The 5946 actual non-defaults predicted as defaults would also have implications in the form of higher monitoring costs, or financial institutions might reject giving credit to borrowers who are unlikely to default.

### 7.2 Backward Elimination Model

```{r ROC stepmodel is}
res_stepmodel <- predict(bwd.model, traindata, type = "response")

ROCpred <- ROCR::prediction(res_stepmodel, traindata$default)

ROCperf <- ROCR::performance(ROCpred, "tpr", "fpr")

plot(ROCperf, 
     colorize = T, 
     print.cutoffs.at = seq(0, by = 0.01),
     text.adj = c(-0.2,1))
```

```{r conMat stepmodel is}
# Use 0.17
pred.stepmodel <- as.factor(ifelse(res_stepmodel > 0.17, "1", "0"))

caret::confusionMatrix(traindata$default,
                       pred.stepmodel,
                       dnn = c("Actual", "Predicted"))
```













## 8 Evaluation of Models Out-of-Sample

### 8.1 Full Model

```{r conMat fullmodel oos}
# Change to testdata instead for out-of-sample evaluation
res_fullmodel <- predict(full.model, testdata, type = "response")

# Use the same threshold as in-sample evaluation
pred.fullmodel <- as.factor(ifelse(res_fullmodel > 0.17, "1", "0"))

caret::confusionMatrix(testdata$default,
                       pred.fullmodel,
                       dnn = c("Actual", "Predicted"))
```

Using the out-of-sample data, the `fullmodel` incorrectly predicted 385 actual defaults as non-default. However, in terms of the accuracy of the model, it was the same as using in-sample data. Since this is out-of-sample data, we could say that banks might turn away 1514 customers as the model predicted that they will default, although they would not.

### 8.2 Backward Elimination Model

```{r conMat stepmodel oos}
res_stepmodel <- predict(bwd.model, testdata, type = "response")

pred.stepmodel <- as.factor(ifelse(res_stepmodel > 0.17, "1", "0"))

caret::confusionMatrix(testdata$default,
                       pred.stepmodel,
                       dnn = c("Actual", "Predicted"))
```













## 9 Final Remarks

Real-world models would be more complicated than these, and may take into account qualitative variables as well. However, this give a baseline to how well logistic regression models would work in binary classification problems. It is important to note that the models in this project do not consider a cost to misclassification, which is an important factor in real-world problems. There could also be a trade-off involved in the costs of false positives and false negatives and there would be a need to weigh these costs to select the best model to be implemented.

## References

<https://bookdown.org/tpinto_home/Regularisation/lasso-regression.html>
