---
title: "Classification Using Logistic Regression Models"
output: 
  github_document:
    toc: true
---

# 1 Introduction

The purpose of this project was to apply what I had learnt about logistic regression and classification in machine learning (ML). It documents (1) the packages and functions that are useful for logistic classification and ML, and (2) the steps to building a logistic classification model in R.

The project has three main parts. In the first part, I introduced what logistic regression is and why it is used in classification in machine learning. In the second part, I provided examples of logistic regression in binary classification (i.e. there are only two outcomes) using an in-built dataset in R. In the third part, I attempted to use logistic regression in multinomial classification (i.e. there are more than two outcomes) using an in-built dataset as well.

(Intro to be further refined at the end)

## 1.1 What is logistic regression and why is it used in classification problems?

## 1.2 What are the assumptions needed for logistic regression?

This sub-section lists the common assumptions of logistic regressions and some of the ways to check them.

### 1.2a Type of logistic regression matches the response variable

The type of logit regression needs to match the type of outcome. In most cases, logit regression is used for binary outcomes but there are cases where it might be used for multinomial or ordinal outcomes and appropriate models need to be used for each type.

**To check assumption**: Check the number of unique outcomes/response variable. If there are only two outcomes, it is very likely to be a binary classification problem. If there are more than two outcomes, we are dealing with multinomial or ordinal classification problems.

### 1.2b Observations are independent

Observations need to be explanatory of one another, and not be affected by other observations

**To check assumption**: Plot the residuals against order of observations and check if there is a random pattern. If pattern is not random, it indicates possibility of correlation between observations. However, autocorrelation should be less of a concern with cross-sectional data as long as the design of the study ensures that data are collected from random samples and there are no repeated or paired observations.

### 1.2c No multicollinearity among explanatory variables

Independent variables should not be highly correlated with one another as it affects the variance of estimated coefficients and statistical significance of estimated coefficients becomes unreliable.

**To check assumption**: A simple method would be to use a correlation matrix to find highly correlated explanatory variables. Another method is to calculate the variance inflation factor (VIF) of each explanatory variable. A common rule of thumb is that VIF more than 10 indicates problem of multicollinearity.

### 1.2d Linear relationship between continuous explanatory variables and logit of response variable

Continuous explanatory variables and the logit of the response variable (or log-odds) have a linear relationship. Logit regression does not require linearity between continuous explanatory variables and response variable.

**To check assumption**: Plot log-odds against each continuous explanatory variable or use the Box-Tidwell test.

### 1.2e No highly influential outliers

There should be no influential outliers that could affect the outcome of the model. Removing such observations are possible if they are determined to be incorrectly entered or measured. It is also possible to compare the regression results with and without the outliers, and note how the results differ.

**To check assumption**: Calculate Cook's Distance to find influential observations and standardized residuals to find outliers.

### 1.2f Large sample size

A large sample size is needed to produce conclusive results.

# 2 Packages Required

```{r load packages, message=FALSE}
library(car) 
library(caret) # For building ML models
library(corrplot) # For correlation plot
library(tidyverse) # For ggplot and dplyr packages
library(mlbench) # For data sets
```

These packages provide the necessary functions and data sets that can be used for learning and building logistic classification models.

# 3 Basics of Logistic Regression

To start off with the basics of logistic regression in R, it is best to learn how it works with binary outcomes first. Such outcomes could be 0 or 1, Yes or No, Male or Female, etc. with only 2 possible outcomes. I selected the Pima Indians Diabetes Database in the **`mlbench`** package for this example.

## 3.1 Exploratory Data Analysis

Let us first load the data and perform an Exploratory Data Analysis before the modelling task.

```{r load PimaIndiansDiabetes dataset}
data("PimaIndiansDiabetes")
```

```{r structure and summary}
str(PimaIndiansDiabetes)

summary(PimaIndiansDiabetes)
```

```{r head and tail of df}
head(PimaIndiansDiabetes, n = 4)

tail(PimaIndiansDiabetes, n = 4)
```

By entering `??PimaIndiansDiabetes` into the console, we can get a description of the variables in the data frame. I have listed the variables and its description in the table below.

| Variable | Description                                           |
|----------|-------------------------------------------------------|
| pregnant | Number of times pregnant                              |
| glucose  | Plasma glucose concentration (glucose tolerance test) |
| pressure | Diastolic blood pressure (mm Hg)                      |
| triceps  | Triceps skin fold thickness (mm)                      |
| insulin  | 2-Hour serum insulin (mu U/ml)                        |
| mass     | Body mass index (weight in kg/(height in m)^2)        |
| pedigree | Diabetes pedigree function                            |
| age      | Age (years)                                           |
| diabetes | Class variable (test for diabetes)                    |

The **`str`** function shows that there are 768 observations with 9 variables and only the variable *`diabetes`* is a factor while all other variables are numeric.

The **`summary`** function reveals that there are possibly some errors in the data, with value of zero in variables *`glucose`*, *`pressure`*, *`triceps`*, and *`mass`*. I would remove these observations to keep things simple.

```{r filter out obs}
data <- PimaIndiansDiabetes %>%
  dplyr::filter(glucose != 0 & pressure != 0 & triceps != 0 & mass != 0)

summary(data)
```

It should be noted that *`insulin`* has zero values as well, which could be due to not receiving insulin despite having diabetes. It could be interesting to look into the distribution of these values. Before that, it might be convenient to split the input and output variables.

```{r split input and output}
input <- data["diabetes"]
summary(input)

output <- subset(data, select = -c(diabetes))
summary(output)
```

Let us take a look at the distribution of the input and output variables in a univariate plot 

```{r dist of diabetic test}
plot(input, main = "Distribution of Diabetic Test",
     ylab = "Frequency")
```













# 4 Logistic Regression for Binary Classification






# 5 Logistic Regression for Multinomial Classification






# 6 Logistic Regression for Ordinal Classification















# References

<https://www.statisticssolutions.com/free-resources/directory-of-statistical-analyses/assumptions-of-logistic-regression/>

<https://towardsdatascience.com/assumptions-of-logistic-regression-clearly-explained-44d85a22b290>

<https://www.statology.org/assumptions-of-logistic-regression/>

<https://www.theanalysisfactor.com/outliers-to-drop-or-not-to-drop/>
